Comparison of approaches to using machine learning models to increase the efficiency of the global search algorithm for solving multicriterial problems

### 2024-06-25 2024/OPTIMA/OPTIMA_2024_MCO.tex
add some references

For solving the multiextremal problems the more effective algorithms are used which can be divided into two key classes, namely, into deterministic and meta-heuristic (nature-inspired) methods \cite{ML_MCO_2023,Evtushenko2014,Deb2002,Durillo2010,Mostaghim2007,NDG09,RC05,ZLT01,Gergel2018,GergelKozinov2020}.

@@

In the publication \cite{ML_MCO_2023} we proposed an approach to solving multicriterial optimization problems combining the classic information-statistical global search algorithm with machine learning methods. In the framework of the approach a problem of multicriterial optimization was transformed to a series of scalar optimization subproblems. In the course of solving scalar subproblems, on the base of machine learning methods a hyperplane was built and gradually refined which separated points of effective and ineffective solutions in the criteria space. Additional information connected with the separating hyperplane was embedded in the computational scheme of the optimization algorithm and, as a result, the number of trials (criteria evaluations) in domains with ineffective solutions was significantly decreased. The experimental results showed that in comparison with a number of heuristic algorithms, the search for effective solutions was several times faster. At the same time, this approach can be improved because utilizing only the separating plane limits the class of machine learning techniques which can be applied for enhancing the global search algorithm.

>>>>

For solving the multiextremal problems the more effective algorithms are used which can be divided into two key classes, namely, into deterministic (see, e.g., \cite{Evtushenko2014,Gergel2018,GergelKozinov2020,Paulavicius2020,Jones2021}) and meta-heuristic (see, e.g., \cite{Battiti2009,Gendreau2010,Eiben2015}) methods .

@@

In the paper \cite{ML_MCO_2023} we proposed an approach to solving multicriterial optimization problems combining the classic information-statistical global search algorithm with machine learning methods. In the framework of the approach a problem of multicriterial optimization was transformed to a series of scalar optimization subproblems. In the course of solving scalar subproblems, on the base of machine learning methods a hyperplane was built and gradually refined which separated points of effective and ineffective solutions in the criteria space. Additional information connected with the separating hyperplane was embedded in the computational scheme of the optimization algorithm and, as a result, the number of trials (criteria evaluations) in domains with ineffective solutions was significantly decreased. The experimental results showed that in comparison with a number of heuristic algorithms, the search for effective solutions was several times faster. At the same time, this approach can be improved because utilizing only the separating plane limits the class of machine learning techniques which can be applied for enhancing the global search algorithm.
